---
title: "Recovering the table behind a published kappa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recovering the table behind a published kappa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(enum2x2)
```

A study comparing two binary criteria on one population computes its numbers from a
2x2 table, then publishes the numbers and not the table. The quantity a reader wants
--- how many cases the two criteria classify differently, and in which direction ---
is gone. It is usually still recoverable, because the same reports print the sample
size and both marginal totals, and those pin the table down.

This vignette walks three cases: one the published digits determine exactly, one where
a coefficient of similar size belongs to an entirely different joint classification,
and one where the published figures admit no table at all.

## Cell names

Cell names are those of `metafor::conv.2x2`.

|            | criterion 2 + | criterion 2 - | total |
|------------|--------------:|--------------:|------:|
| criterion 1 + | `ai`       | `bi`          | `n1i` |
| criterion 1 - | `ci`       | `di`          |       |
| total         | `n2i`      |               | `ni`  |

## A determined case

A pooled series of 768 patients, two readings of the DSM-5 delirium criteria,
published at 60% raw agreement and kappa = 0.29. The strict reading is positive in 158
patients and the relaxed reading in 466. Figures are passed as the strings the source
printed, because `"0.29"` and `"0.290"` imply different intervals and a double cannot
tell them apart.

```{r delirium}
delirium <- recover2x2(768, n1i = 158, n2i = 466, kappa = "0.29",
                       agreement = "60", agreement.as.percent = TRUE)
delirium
```

One table survives, so the report determines its own 2x2 exactly:

```{r delirium-table}
tab <- unique_table(delirium)
tab
matrix(c(tab$ai, tab$bi, tab$ci, tab$di), nrow = 2, byrow = TRUE,
       dimnames = list(c("strict +", "strict -"), c("relaxed +", "relaxed -")))
```

The two readings never cross-classify. No patient is strict-positive and
relaxed-negative, and every one of the patients they disagree on falls the same way:

```{r delirium-direction}
c(strict_only = tab$bi, relaxed_only = tab$ci)
asymmetry2x2(tab$bi, tab$ci)
```

An asymmetry of 1 means the disagreement runs entirely in one direction. One reading is
simply wider than the other, and its positives contain the other's. The observed
agreement is also the highest these two marginals permit, so 0.29 is the ceiling here
rather than a shortfall from it:

```{r delirium-ceiling}
c(published = kappa2x2(tab$ai, tab$bi, tab$ci, tab$di),
  ceiling = kappa_max(158 / 768, 466 / 768))
```

None of that is in "kappa = 0.29", and kappa = 0.29 is what the paper printed.

## A coefficient of similar size, a different structure

A second comparison from the same corpus: 20,306 cases, the first criterion positive in
866 and the second in 1,603, published at kappa = 0.22. The coefficient is close to the
first one. The joint classification is not.

```{r cohort}
cohort <- recover2x2(20306, n1i = 866, n2i = 1603, kappa = "0.22")
cohort
```

Two decimals on a sample this size no longer determine the table, so the answer is a set
of `r n_tables(cohort)` tables and the report is set-identified rather than unique. The
span of each cell is what the printed digits leave open:

```{r cohort-ranges}
cell_ranges(cohort)
```

Every table in the set has substantial counts in both off-diagonal cells, so here the
two criteria do cross-classify:

```{r cohort-asymmetry}
asym <- asymmetry2x2(cohort$tables$bi, cohort$tables$ci)
range(asym)
```

```{r cohort-plot, fig.width = 6, fig.height = 4, fig.alt = "Each compatible table plotted by its two discordant counts"}
plot(cohort$tables$bi, cohort$tables$ci, pch = 19,
     xlab = "criterion 1 only (bi)", ylab = "criterion 2 only (ci)",
     main = "The tables kappa = 0.22 admits on 20,306 cases")
```

Set side by side, the two reports differ in the only thing a reader would use them for:

```{r contrast}
data.frame(
  case = c("delirium, 768", "cohort, 20306"),
  kappa = c("0.29", "0.22"),
  tables = c(n_tables(delirium), n_tables(cohort)),
  discordant_one_way = c(tab$bi, paste(range(cohort$tables$bi), collapse = "-")),
  discordant_other_way = c(tab$ci, paste(range(cohort$tables$ci), collapse = "-")),
  asymmetry = c(round(asymmetry2x2(tab$bi, tab$ci), 2),
                paste(round(range(asym), 2), collapse = "-")))
```

A published kappa near 0.2 to 0.3 is consistent with two criteria that never disagree in
one direction, and with two criteria whose disagreement is split three to one. The
coefficient does not distinguish them; the recovered table does.

## A case with no table behind it

A third report: 370 patients, the first criterion positive in 165 and the second in 160,
published at kappa = 0.48 and 73% agreement.

```{r infeasible}
bad <- recover2x2(370, n1i = 165, n2i = 160, kappa = "0.48",
                  agreement = "73", agreement.as.percent = TRUE)
bad$status
bad$reason
```

The status is `infeasible`, and it is kept apart from `insufficient`, which is what a
report gets when a required figure was never published at all:

```{r insufficient}
recover2x2(370, n1i = 165, n2i = 160)$status
```

The reason names the figure that excluded. Dropping the published agreement leaves a
table, so it is the agreement and the kappa that fail to cohere rather than the kappa
and the marginals:

```{r infeasible-without}
unique_table(recover2x2(370, n1i = 165, n2i = 160, kappa = "0.48"))
```

What this licenses is narrow. It says the published figures admit no common table under
the definitions and rounding conventions used here. It does not say an error was made: a
denominator, an analysis set, or a version of a variable may simply have gone unstated,
and a weighted kappa or a different handling of missing cases would give different
numbers from the same study. The finding is that the printed figures cannot all be read
off one 2x2 --- not that any of them is wrong.

## A series of comparisons

`enum.2x2` takes a data frame, in the shape `metafor::conv.2x2` uses.

```{r frame}
dat <- data.frame(
  study = c("delirium pooled", "cohort", "third report"),
  ni    = c(768, 20306, 370),
  n1i   = c(158, 866, 165),
  n2i   = c(466, 1603, 160),
  kappa = c("0.29", "0.22", "0.48"))

out <- enum.2x2(kappa = kappa, ni = ni, n1i = n1i, n2i = n2i, data = dat)
out[, c("study", "status", "ntables", "ai", "ai.min", "ai.max")]
```

Where `conv.2x2` appends four columns of reconstructed counts, this appends the same four
and fills them only where the printed digits determine them, alongside the span of each
cell and the number of compatible tables. Row 2 has `ai` missing and a span of
`r out$ai.min[2]` to `r out$ai.max[2]`, which is the information a point estimate would
have replaced with a single number.

The full set for any row remains reachable:

```{r frame-recoveries}
attr(out, "recoveries")[[2]]
```

## Why rounding is the whole problem

With exact inputs the algebra is one line. For positive rates `p1` and `p2` on `ni`
observations, expected agreement is `p1 * p2 + (1 - p1) * (1 - p2)`, observed agreement
is `kappa * (1 - p_e) + p_e`, and the both-positive cell is
`(p_o + p1 + p2 - 1) / 2`. Nobody prints exact inputs. A marginal printed `4.26%` and a
kappa printed `0.22` are intervals:

```{r intervals}
rounding_interval("0.22")
rounding_interval("4.26", as.percent = TRUE)
counts_rounding_to("4.26", 20306, as.percent = TRUE)
```

Membership is decided on the exact form of those intervals, against a table's statistic
written as a ratio of integers, so both sides are rational and the comparison carries no
tolerance:

```{r exact}
exact_interval("0.22")
exact_kappa(158, 0, 308, 302)
```

The interval is closed at both ends. A value falling exactly on a boundary rounds up
under one convention and down under another, and sources do not say which they used. The
table `(2, 0, 2, 20)` on 24 cases has kappa exactly 5/8, which a source prints as `"0.62"`
under one rule and `"0.63"` under the other, and the table survives either printing:

```{r endpoint}
exact_kappa(2, 0, 2, 20)
unique_table(recover2x2(24, n1i = 2, n2i = 4, kappa = "0.62"))
unique_table(recover2x2(24, n1i = 2, n2i = 4, kappa = "0.63"))
```

## What it will not do

It does not say which criterion is correct; neither does the table. It does not separate
disagreement *between* criteria from inconsistent *application* of either, which would
need the same criterion applied twice to the same cases. Three or more categories are out
of scope, since the identity above is the binary case.
